A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-06 tidyindex.qmd
a software design paragraph and some input/output for temporal processing

## Pipeline steps for constructing indexes

@@

@@

\begin{equation}
f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
\end{equation}

where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$. 
A typical example of temporal processing is aggregation, which is used in the drought index SPI to measure the lack of precipitation for meteorological drought.
In SPI, monthly precipitation is aggregated by a time scale parameter $k$: $x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j),$ where $j^\prime$ is the new time index after the aggregation. In this notation, each spatial location is separately aggregated and precipitation is summed from $k$ month back, $j^\prime - k + 1$, to the current period, $j^\prime$, to create the aggregated series, indexed by $j^\prime$.

*more explicit on k will influence 1) long term vs. short term, 2) uncertainty*

The choice of time scales parameter $k$ can result in variation in the calculated index values: a small $k$ of 3 or 6 months produces the index more sensitive to individual months, while a large $k$ of 24 or 36, an equivalent to a 2- or 3-year aggregation, gives dryness information relative to the long term condition. As will be shown in section [SECTION EXAMPLE], this variation may even lead to conflicting conclusions on the dry/wet condition of the area, highlighting the importance to account for index uncertainty when interpreting index values for decision-making.

*Effective drought index*

@@

@@

>>>>

## Pipeline steps for constructing indexes {#sec-pipeline-steps}

@@

**Input: ** multivariate data with temporal coordinates

**Output: ** multivariate data with aggregated temporal axis

@@

\begin{equation}
x(\mathbf{s};\mathbf{t}^\prime)  f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
\end{equation}

where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$. 
A typical example of temporal processing is aggregation and in the drought community, these indexes are often called multi-scalar indexes.  For example,  Standardized Precipitation Index (SPI) aggregates monthly precipitation using a time scale parameter $k$ in the form of

$$x(s_i;t_{j^\prime}) = \sum_{j = j^\prime}^{j^\prime + k - 1}x(s_i; t_j),$$

where $j^\prime$ is the new time index after the aggregation. Multi-scalar drought indexes are widely used to study different types of drought events. A smaller value of the time scales parameter $k$ is used for short term drought events, while a large $k$ of 18 or 24, an equivalent to a 1.5- or 2-year aggregation, is often used for long term assessment.

@@

**Input: ** Multivariate data with coordinates

**Output: ** Multivariate data with a new set of coordinates

@@

Tidyindex, as suggested by its name, follows the style and design principles of the tidyverse suite. The package is designed to simplify index construction for human analysts, ensure consistency across functions, and enable composability across modules to construct indexes. With the package, analysts can pass a data object through a set of pipeline modules introduced in @sec-pipeline-steps, offering different operations as standalone functions. For example, the dimension reduction module includes linear weighting, geometrical weighting, and manual expressions as possible methods to combine multivariate data. These methods are initially evaluated as a recipe of the operation without the data, and then, at the pipeline level, evaluating with the data using tidy evaluation, with the package `rlang` heavily used throughout the package.

### 2023-08-06 tidyindex.tex
a software design paragraph and some input/output for temporal processing

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\hypertarget{pipeline-steps-for-constructing-indexes}{%
\subsection{Pipeline steps for constructing
indexes}\label{pipeline-steps-for-constructing-indexes}}

@@

@@

\begin{equation}
f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
\end{equation}

where \(\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}\) is the
parameters associated with the temporal operation and \(d_{\psi}\) is
the number of parameter of \(\psi\). A typical example of temporal
processing is aggregation, which is used in the drought index SPI to
measure the lack of precipitation for meteorological drought. In SPI,
monthly precipitation is aggregated by a time scale parameter \(k\):
\(x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j),\)
where \(j^\prime\) is the new time index after the aggregation. In this
notation, each spatial location is separately aggregated and
precipitation is summed from \(k\) month back, \(j^\prime - k + 1\), to
the current period, \(j^\prime\), to create the aggregated series,
indexed by \(j^\prime\).

\emph{more explicit on k will influence 1) long term vs.~short term, 2)
uncertainty}

The choice of time scales parameter \(k\) can result in variation in the
calculated index values: a small \(k\) of 3 or 6 months produces the
index more sensitive to individual months, while a large \(k\) of 24 or
36, an equivalent to a 2- or 3-year aggregation, gives dryness
information relative to the long term condition. As will be shown in
section {[}SECTION EXAMPLE{]}, this variation may even lead to
conflicting conclusions on the dry/wet condition of the area,
highlighting the importance to account for index uncertainty when
interpreting index values for decision-making.

\emph{Effective drought index}

@@

@@

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, breakable, boxrule=0pt, frame hidden, enhanced]}{\end{tcolorbox}}\fi

@@

\hypertarget{sec-pipeline-steps}{%
\subsection{Pipeline steps for constructing
indexes}\label{sec-pipeline-steps}}

@@

\textbf{Input: } multivariate data with temporal coordinates

\textbf{Output: } multivariate data with aggregated temporal axis

@@

\begin{equation}
x(\mathbf{s};\mathbf{t}^\prime)  f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
\end{equation}

where \(\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}\) is the
parameters associated with the temporal operation and \(d_{\psi}\) is
the number of parameter of \(\psi\). A typical example of temporal
processing is aggregation and in the drought community, these indexes
are often called multi-scalar indexes. For example, Standardized
Precipitation Index (SPI) aggregates monthly precipitation using a time
scale parameter \(k\) in the form of

\[x(s_i;t_{j^\prime}) = \sum_{j = j^\prime}^{j^\prime + k - 1}x(s_i; t_j),\]

where \(j^\prime\) is the new time index after the aggregation.
Multi-scalar drought indexes are widely used to study different types of
drought events. A smaller value of the time scales parameter \(k\) is
used for short term drought events, while a large \(k\) of 18 or 24, an
equivalent to a 1.5- or 2-year aggregation, is often used for long term
assessment.

@@

\textbf{Input: } Multivariate data with coordinates

\textbf{Output: } Multivariate data with a new set of coordinates

@@

Tidyindex, as suggested by its name, follows the style and design
principles of the tidyverse suite. The package is designed to simplify
index construction for human analysts, ensure consistency across
functions, and enable composability across modules to construct indexes.
With the package, analysts can pass a data object through a set of
pipeline modules introduced in Section~\ref{sec-pipeline-steps},
offering different operations as standalone functions. For example, the
dimension reduction module includes linear weighting, geometrical
weighting, and manual expressions as possible methods to combine
multivariate data. These methods are initially evaluated as a recipe of
the operation without the data, and then, at the pipeline level,
evaluating with the data using tidy evaluation, with the package
\texttt{rlang} heavily used throughout the package.
